---
title: DANL Project
subtitle: "Investing Strategies: Knowledge for the Short and Long Term"
author: 
  - Name 1:   Christopher Taratko
  - Name 2:   Jack Katz
  - Name 3:   Aaron Zalen
  - Name 4:   John Christanis
format: 
  html: 
    code-fold: true

toc: true
toc-title: "Table of Contents"
toc-depth: 2
number-sections: true

fig-width: 9

execute:
  code-tools: true
  message: false
  warning: false

from: markdown+emoji
---

```{r setup}
#| include: false


library(tidyverse)
library(skimr)
library(ggthemes)
library(hrbrthemes)
library(corrplot)
library(ggplot2)


StocksUpdated2 <- read_csv(
  "/Users/christophertaratko/ProjectsList/Data Analytics/Personal Website/ctaratko.github.io/StocksUpdated2.csv")


Volatility_analysis <- StocksUpdated2 %>% 
  separate(Date, into = c("Month/Day","Year"), sep = 6 )

```

# Introduction

The goal intended of the research and analysis of the stocks performed was to help people understand what certain stocks would classify as, whether it be a good investment overall, something preferable for the short/long run, what the volatility of a stock is, and other performance functions that would help the reader grasp what to look for when investing.

# Data

We retrieved the dataset "Stock Market: Historical Data of Top 10 Companies" from Kaggle (<https://www.kaggle.com/datasets/khushipitroda/stock-market-historical-data-of-top-10-companies>). The data includes 7 variables and 25160 observations. The variables are Company which is a factor variable for each company name. The data.frame is `Stocks Updated` and is

## Summary Statistics

```{r}
summary(StocksUpdated2)
```

# Jack Katz - Research Question 1: Which stocks have the most historical volatility.

### Creating new variables for analysis, Separating by month day and year will make it easier to group and filter observations by year.

## Historical volatility of each company for each year.
A simple measure of historical volatility is calculated by finding the standard deviation of the close prices over a time series. In this case we found the standard deviation for each stock for each year and then showed this data on ggplot.

```{r}
Historical_Volatility <- Volatility_analysis %>% 
  group_by(Company, Year) %>% 
  summarise(Volatility = sd(`Close/Last`))

Volatility <- ggplot(Historical_Volatility,
       mapping = aes(x = as.numeric(Year),
                     y = Volatility
                     )) +
  geom_point(color = "blue") +
  geom_line(color = "green") +
  facet_wrap( . ~Company) +
  ggtitle("Stock Price Volatility from 2013-2023", subtitle = "In dollars") +
  labs(x = "Year") +
  theme_economist_white()

Volatility + scale_x_continuous(breaks = scales::breaks_width(2))

```

# Christopher Taratko - Which Stocks are Better Short Term Investments vs Long Term Investments?

### Figuring out which stocks could be used at Short and Long term investments was through the use of their PVT Indicator Values & the MAD (Mean Absolute Deviation) of their Closing Value Change.

```{r}
StocksChange <- StocksUpdated2 %>% 
  select(Company, Date, `Close/Last`, Volume) %>% 
  group_by(Company) %>% 
  mutate(CL_Chg = `Close/Last` - lag(`Close/Last`),
         V_Chg = Volume - lag(Volume),
         PVT = (CL_Chg / V_Chg))

ggplot(StocksChange, aes( x = Company, y = PVT)) +
  geom_hex()

# PVT = Price Volume Trend Indicator. ( ∆ in Price / ∆ in Volume )

```

```{r}
# Object #2:



StocksManip <- StocksUpdated2 %>% 
  mutate(CL_Change =`Close/Last` - lag(`Close/Last`))

StocksManip <- StocksManip %>% 
  select(Company,Date,CL_Change)

StocksManip <- StocksManip %>% 
  group_by(Company) %>% 
  arrange(Date) %>% 
  pivot_wider(names_from = Company, values_from = CL_Change)

summary(StocksManip)

StocksMAD <- StocksManip %>% 
  mutate(CSCO_mad = CSCO - mean(CSCO, na.rm = T),
         QCOM_mad = QCOM - mean(QCOM, na.rm = T),
         META_mad = META- mean(META, na.rm = T),
         AMZN_mad = AMZN - mean(AMZN, na.rm = T),
         TSLA_mad = TSLA - mean(TSLA, na.rm = T),
         AMD_mad = AMD - mean(AMD, na.rm = T),
         NFLX_mad = NFLX - mean(NFLX, na.rm = T),
         AAPL_mad = AAPL - mean(AAPL, na.rm = T),
         SBUX_mad = SBUX - mean(SBUX, na.rm = T),
         MSFT_mad = MSFT - mean(MSFT, na.rm = T))




mean(StocksManip$AAPL)
mean(StocksManip$MSFT)


StocksIsolated <- StocksMAD %>% 
  select(ends_with("_mad"))

Isolated_long <- StocksIsolated %>%
  pivot_longer(cols = ends_with("_MAD"),names_to = "Company", values_to = "MAD")


ggplot(Isolated_long, aes(y = Company, x = MAD)) +
  geom_hex()
```

# Aaron Zalen - Does higher volume correlate with a larger increase in price?

```{r}
# Data manipulation -----------------------------------------------------------------

#Opening File

#First we change the values to be numerical 
stocks <- StocksUpdated2 %>% 
  mutate(Change=`Close/Last`-Open) %>% 
  mutate(PercentageChange = ((`Close/Last` - Open) / Open) * 100) %>% 
  group_by(Company)


# Graph 1 -----------------------------------------------------------------

ggplot(stocks, aes(y = PercentageChange, x = log10(Volume))) +
  geom_point(alpha = 0.1, color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  facet_wrap(~ Company, scales="free_y") +
  labs(
    title = "Percentage Change vs. Volume",
    x = "Log10 of Volume",
    y = "Percentage Change",
    caption = "Blue points represent data, Red line represents linear fit"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    strip.text = element_text(size = 12),
    legend.position = "none"  # Remove legend for better clarity
  )

```
```{r}
# Different graph --------------------------------------------------------------

correlation_by_group <- stocks %>%
  group_by(Company) %>%
  summarise(correlation = cor(PercentageChange, Volume))

ggplot(correlation_by_group, aes(x=Company, y=correlation))+
  geom_col()+
  labs(title= "How strongly does Volume and Percentage Change in Price Correlate?")

```

### The change in volume does not appear to have any effect on the stock's price

# John Christanis - How are the Individual Stocks Correlated?
```{r}
StocksJ <- StocksUpdated2 %>% 
  mutate(Change=`Close/Last`-Open) %>% 
  mutate(PercentageChange = ((`Close/Last` - Open) / Open) * 100)

StockJ <- StocksJ %>% 
  select(Company|PercentageChange|Date) %>% 
  pivot_wider(names_from = "Company", values_from = "PercentageChange")


correlation_matrix <- cor(StockJ[, sapply(StockJ, is.numeric)], use = "pairwise.complete.obs")

print(correlation_matrix)

corrplot(correlation_matrix, method = "circle", type = "upper", tl.col = "black", tl.srt = 45)

```

# Conclusion:
