---
title: DANL Project
subtitle: "Investing Strategies: Knowledge for the Short and Long Term"
author: 
  - Name 1:   Christopher Taratko
  - Name 2:   Jack Katz
  - Name 3:   Aaron Zalen
  - Name 4:   John Christanis

toc: true
toc-title: "Table of Contents"
toc-depth: 2
number-sections: true

fig-width: 9

execute:
  code-tools: true
  message: false
  warning: false

from: markdown+emoji
---

```{r setup}
#| include: false
library(tidyverse)
library(skimr)
library(ggthemes)
library(hrbrthemes)

theme_set(theme_ipsum()+
          theme(strip.background =element_rect(fill="lightgray"),
                axis.title.x = 
                  element_text(angle = 0,
                               size = rel(1.33),
                               margin = margin(10,0,0,0)),
                axis.title.y = 
                  element_text(angle = 0,
                               size = rel(1.33),
                               margin = margin(0,10,0,0))
                )
          )
```

# Introduction

The goal intended of the research and analysis of the stocks performed was to help people understand what certain stocks would classify as, whether it be a good investment overall, something preferable for the short/long run, what the volatility of a stock is, and other performance functions that would help the reader grasp what to look for when investing.

```{r}
1 + 1
```

# Data

The data set comes `data` comes from kaggle.com from July 18th, 2023, where the code explores a variety of stocks from www.nasdaq.com to give you an easy comparison, but is still fairly broad with what they give you

## Summary Statistics

```{r}
mpg <- ggplot2::mpg
```

```{r}
#| results: asis
#| echo: false
rmarkdown::paged_table(mpg) 
```

```{r}
skim(mpg) %>% 
  select(-n_missing)
```

## MPG and a Type of Cars

The following boxplot shows how the distribution of highway MPG (`hwy`) varies by a type of cars (`class`) :blue_car: :truck: :minibus:.

```{r}
ggplot(data = mpg) +
  geom_boxplot(aes(x = class, y = hwy, fill = class),
               show.legend = F) +
  labs(x = "Class", y = "Highway\nMPG") 
```
