---
title: "Capstone Paper: Household Financial Health"
author: "Christopher Taratko"
format:
  html:
    backgroundcolor: "#fdf6e3"
toc: true
---

```{r, include=FALSE}
library(rmarkdown)
library(tidyverse)
library(broom)
library(stargazer)
library(skimr)

library(margins)
library(yardstick)
library(WVPlots)
library(pROC)

library(glmnet)
library(Matrix)

library(rmarkdown)
library(hrbrthemes)
library(ggthemes)

library(plm)
library(zoo)
setwd("/Users/christophertaratko/ProjectsList/Data Analytics/Personal Website/ctaratko.github.io/Data/CapstoneDatasets")
model_prep <- read_csv("Housing.csv")
# setwd("/Users/christophertaratko/ProjectsList/Data Analytics/Personal Website/ctaratko.github.io/Spring2026Projects")

my_theme <- theme(
    plot.title = element_text(family = "Times New Roman", color = "#700000", face = 'bold'),
    plot.subtitle = element_text(
      face = "italic",
      family = "Times New Roman",
      size = 10
    ),
    axis.title = element_text(family = "Times New Roman", face = "bold"),
    legend.position = "right",
    legend.title = element_text(family = "Times New Roman", face = "bold"),
    legend.text = element_text(family = "Times New Roman", face = "italic"),
    plot.caption = element_text(family = "Times New Roman", size = 7),
    strip.background = element_rect(
      fill = "#fdf6e3",
      color = "black", 
      linewidth = .5
    ),
    
    strip.text = element_text(face = "bold")
  )
```

# Introduction

Current households face the question of whether or not they should spend money to build up their assets, more specifically their long term assets. Long Term Assets is a broad classification of assets, but when specified under a Household, it normally pertains to assets such as:

- Houses

- Car

- Small Businesses

the US consumer body as a whole is at a state where its becoming less and less common to have a larger amount of debt relative to one's income, as as financially sound this is, it also creates concern for future innovation. Entrepreneurship requires risks to be taken, and the US has been the global powerhouse for innovation and research for new products and services, and the US government has backed this innovation with the amount of debt that's been acquired.

As of recently, it is able to be seen that individuals have began choosing new methods of acquiring debt, but for less innovative or valued purposes, but instead for financially illiterate purposes. It is common for people to overspend on their line(s) of revolving credit (credit cards) that have balances due at the end of each month, or else they'll be charged interest on the balance. Other options are people will over-spend on their debit card, and end up utilizing methods such as "Buy-Now-Pay-Later", or "Pay-Day" Loans.

## Thesis Question

- What levels of DTI Ratios (Debt-to-Income) and HPI conditions have been sustainably manageable for the U.S. households based on the ML Outputs?

  - What factors best represent the overall health of Household finances in the United States?

  - How have US financial factors changed over time, and what patterns have emerged from the changes?

# Background and Related Literature

Throughout the process of learning about the data itself, key ideas that were needed to be known going into the research and understanding of this financial data is the connection between financial instruments and the housing market as a whole. The article "House Prices, Home Equity-Based Borrowing, and the US Household Leverage Crisis (*Mian, A, & Sufi, A, 2011*), is a critical analysis of the 2008 Housing Crisis and the ties to household valuations. From this paper, one of the key concepts is how the housing boom didn't only raise household values, but rather encouraged heavy borrowing against rising home equity. This was one of the large factors and and a signficant catalyst for why the collapse of the housing market was so burdening to the general public. Using detailed individual-level credit data from 1997 to 2008, Mian and Sufi demonstrate that households actively extracted equity from their homes during periods of rising housing prices. Their findings show that, on average, households borrowed approximately \$0.25 for every \$1.00 increase in home equity. This behavior significantly contributed to the overall rise in household leverage during the early 2000s.

Importantly, the paper highlights that this borrowing was not primarily used to pay down existing debt, but rather to finance consumption and other expenditures. As a result, households became increasingly leveraged without improving their financial stability. When housing prices began to decline during the financial crisis, these highly leveraged households faced increased financial strain, leading to widespread defaults and a sharp contraction in economic activity.

This relationship between housing prices, borrowing behavior, and financial vulnerability is central to understanding the dataset used in this analysis. Variables such as Debt-to-Income ratios, housing price indices (HPI), and foreclosure rates are directly connected to the mechanisms described in the literature. By incorporating these variables into a panel data framework, this project aims to capture how changes in housing markets and financial conditions influence household financial health across states and over time.

Overall, the background literature emphasizes that household financial health cannot be understood in isolation. Instead, it must be viewed within the broader context of housing markets, credit systems, and macroeconomic dynamics. This perspective guides the empirical analysis conducted in this project.

# Data

The data within this analysis was constructed through multiple sources, those sources are the following:

- **Federal Reserve Economic Data (FRED)**: Provides macroeconomic indicators such as debt levels, income measures, and foreclosure-related variables.

- **Federal Housing Finance Agency (FHFA)**: Supplies the House Price Index (HPI), which measures changes in housing prices over time.

- **Bureau of Labor Statistics (BLS)**: Provides Consumer Price Index (CPI) data used to measure inflation.

- **New York FED**: Provided Specific Debt amounts, delinquency rates, foreclosure rates, and mortgage payment in relation to income.

### Unit of Observation

The data is a Panel (State-Quarter) based model, used to understand more distinct trends throughout each year in order to get more specific values and time frames. Within this data, there is the following variables:

#### **Input Variables**

- **Debt Value** – \[*DebtValue*\]

- **Median Household Income** – \[*Median_H_Inc*\]

- **Average Adjusted Gross Income** – \[*mean_AGI*\]

- **Minimum Wage** – \[*MinWage*\]

- **Houses Pricing Index** – \[*HPI*\]

- **Mortgage Delinquency Percentage** – \[*Delinquency_pct*\]

- **Average Mortgage PMT Percent** (relative to income) – \[*Mean_Mpct*\]

- **Average State Consumer Prices Index** – \[*mean_CPI*\]

- **Foreclosure** Rate – \[*ForeclosureRate*\]

*Panel-Related Variables*:

- **Time** (Quarter Start Date) – \[*time*\]

- **State** – \[*state*\]

- **Year** – \[*year*\]

#### **Output Variable**

- **Debt-to-Income Ratio** – \[*DTI_ratio*\]

```{r, echo=FALSE}
paged_table(model_prep)
```

These values are vital to understanding household financial health due to their relation to the ability for an individual or family to build wealth, financial stress indicators, Cost of Living by state, Housing Prices by State, and the output of being able to see how much a house can under-go with the stress of debt before issues arise.

## Summary Statistics

### Delinquency Rates

```{r, echo=FALSE, warning=FALSE}
model_prep |> 
  filter(year >= 2015) |>
  ggplot(aes(x = Delinquency_pct)) +
    geom_histogram(bins = 20, fill = "#700", color = "grey40") +
    facet_wrap(~year) +
  theme_solarized() +
  my_theme +
  labs(
    title = "Delinquency Rates Over the Past 10 Years",
    x = "Delinquency (%)",
    y = ""
  )
```

Over the past 10 years, delinquent payments have become less and less of an issue. This is seen as an overall good measurement in terms of individuals taking the risk of over-exerting their finances. These values are returning to pre-2008 Delinquency Rates.

### General Descriptive Stats

```{r, echo=FALSE, warning=FALSE}
summary(model_prep)
```

### Debt-to-Income Ratios Over Time

```{r,warning=FALSE, echo=FALSE}
ggplot(model_prep, aes(y = DTI_ratio, x = time, fill = state)) +
  geom_col(show.legend = F) +
  geom_line(show.legend = F) +
  geom_vline(xintercept = as.Date("2008-01-01"), size = .15) +
  geom_vline(xintercept = as.Date("2020-03-01"), size = .15) +
  facet_wrap(~state, scales = "free_y", ncol = 2) +
  theme_solarized() +
  my_theme
```

```{r}
model_prep |> 
  mutate(state = fct_reorder2(state, time, HPI)) |>
  
  ggplot(aes(x = time, y = HPI, color = state)) +
  geom_smooth(se = F, size = .5, method =  "gam") +
  geom_vline(xintercept = as.Date("2008-01-01"), size = .15) +
  geom_vline(xintercept = as.Date("2020-03-01"), size = .15) +
  labs(
    y = "House Pricing Index",
    x = "Time (in quarters)",
    title = "HPI Trend by State",
    subtitle = "US (2003 - 2023)",
    caption = "Vlines:\n Housing Crisis - 2008\n COVID Pandemic - 2020",
    color = "State"
  ) +
  scale_color_solarized() +
  theme_solarized() +
  theme(
    plot.title = element_text(family = "Times New Roman", color = "#700000", face = 'bold'),
    plot.subtitle = element_text(
      face = "italic",
      family = "Times New Roman",
      size = 10
    ),
    axis.title = element_text(family = "Times New Roman", face = "bold"),
    legend.position = "right",
    legend.title = element_text(family = "Times New Roman", face = "bold"),
    legend.text = element_text(family = "Times New Roman", face = "italic"),
    plot.caption = element_text(family = "Times New Roman", size = 7),
  )
```

```{r}
model_prep |> 
  mutate(state = fct_reorder2(state, time, DTI_ratio)) |>

  ggplot(aes(x = time, y = DTI_ratio, color = state)) +
  geom_line(size = .5) +
  # geom_smooth(size = .25, se = F, method = "lm") +
  geom_vline(xintercept = as.Date("2008-01-01"), size = .15) +
  geom_vline(xintercept = as.Date("2020-03-01"), size = .15) +
  labs(
    title = "Debt-To-Income Trend by State",
    subtitle = "US (2003 - 2023)",
    x = "Time (in quarters)",
    y = "Debt-to-Income Ratio",
    caption = "Vlines:\n Housing Crisis - 2008\n COVID Pandemic - 2020",
    color = "State"
  ) +
  theme_solarized() +
  scale_color_solarized() +
  theme(
    # axis.text.x = element_text(angle = 15),
    axis.text = element_text(family = "Times New Roman"),
    plot.title = element_text(family = "Times New Roman", color = "#700000", face = "bold"),
    plot.subtitle = element_text(
      face = "italic",
      family = "Times New Roman",
      size = 10
    ),
    axis.title = element_text(family = "Times New Roman", face = "bold"),
    legend.position = "right",
    legend.title = element_text(family = "Times New Roman", face = "bold"),
    legend.text = element_text(family = "Times New Roman", face = "italic"),
    plot.caption = element_text(family = "Times New Roman", size = 7),
  )
```

```{r, warning=FALSE, echo=FALSE}
model_prep |> 
  ggplot(aes(x = log(ForeclosureRate), y = DTI_ratio, color = state)) +
  geom_point(alpha = .5, show.legend = F) +
  facet_wrap(~state, ncol = 2) +
  theme_solarized() +
  my_theme +
  labs(
    x = "Log(Foreclosure Rate)",
    y = "Debt-to-Income Ratio"
  )
```

```{r, warning=FALSE, echo=FALSE}
model_prep |> 
  group_by(state) |> 
  arrange(time) |> 
  mutate(
    start_inc = first(Median_H_Inc),
    cum_inflation = cumprod(1 + mean_CPI), 
    state_inflation_line = first(Median_H_Inc) * (mean_CPI / first(mean_CPI))
  ) |> 
  ungroup() |>
  mutate(state = fct_reorder2(state, time, Median_H_Inc)) |>
  ggplot(aes(x = time, y = Median_H_Inc)) +
  geom_line(aes(y = state_inflation_line), 
            color = "black", 
            linetype = "dashed", 
            alpha = 0.6) +
  geom_line(aes(color = state), size = 0.6) +
  facet_wrap(~state, ncol = 2) +
  geom_vline(xintercept = as.Date("2008-01-01"), size = .2, alpha = 0.5) +
  geom_vline(xintercept = as.Date("2020-03-01"), size = .2, alpha = 0.5) +
  scale_y_continuous(labels = scales::dollar) +
  scale_color_solarized() +
  theme_solarized() +
  labs(
    y = "Median Household Income",
    x = "Time (in quarters)",
    title = "Median Household Income vs. 1.5% Inflation (Low End)",
    subtitle = "Dashed line represents standard inflation baseline from 2003",
    caption = "Vlines: 2008 Housing Crisis, 2020 COVID\nDash: 2% Inflation Baseline - www.minneapolisfed.org",
  ) +
  my_theme +
  theme(
    legend.position = "none",
    strip.background = element_rect(fill = "#eee8d5"),
    strip.text = element_text(family = "Times New Roman", face = "bold"),
    text = element_text(family = "Times New Roman"),
    plot.title = element_text(family = "Times New Roman", color = "#700000", face = 'bold'),
  )
```

# Empirical Strategy / Methods

The empirical strategy employs panel data regression techniques using the plm (Panel Linear Model) framework to analyze the determinants of household financial health across U.S. states over time. Panel data allows for controlling both cross-sectional (state-level) and time-series variation, improving estimation accuracy relative to purely cross-sectional or time-series models.

Three model specifications are estimated: Pooled OLS, Fixed Effects (FE), and Two-Way Fixed Effects (TWFE). Each model provides a different level of control for unobserved heterogeneity.

## Pooled OLS Model

$$\begin{aligned}
\log(DTI\_ratio_{it}) =\;&
\beta_1 \log(DebtValue_{it})
+ \beta_2 \log(Median\_H\_Inc_{it})
+ \beta_3 MinWage_{it} \\
&+ \beta_4 \log(HPI_{it})
+ \beta_5 Delinquency\_pct_{it}
+ \beta_6 Med\_Mpct_{it} \\
&+ \beta_7 \log(mean\_CPI_{it})
+ \beta_8 ForeclosureRate_{it}
+ \varepsilon_{it}
\end{aligned}$$

## Fixed Effect Model

$$\begin{aligned}
\log(DTI\_ratio_{it}) =\;&
\beta_1 \log(DebtValue_{it})
+ \beta_2 \log(Median\_H\_Inc_{it})
+ \beta_3 MinWage_{it} \\
&+ \beta_4 \log(HPI_{it})
+ \beta_5 Delinquency\_pct_{it}
+ \beta_6 Med\_Mpct_{it} \\
&+ \beta_7 \log(mean\_CPI_{it})
+ \beta_8 ForeclosureRate_{it}
+ \alpha_i
+ \varepsilon_{it}
\end{aligned}$$

### Interpretation:

- $\alpha_i$: captures all **time-invariant state characteristics**

## Two-Way Fixed Effect Model

$$\begin{aligned}
\log(DTI\_ratio_{it}) =\;&
\beta_1 \log(DebtValue_{it})
+ \beta_2 \log(Median\_H\_Inc_{it})
+ \beta_3 MinWage_{it} \\
&+ \beta_4 \log(HPI_{it})
+ \beta_5 Delinquency\_pct_{it}
+ \beta_6 Med\_Mpct_{it} \\
&+ \beta_7 \log(mean\_CPI_{it})
+ \beta_8 ForeclosureRate_{it}
+ \alpha_i
+ \gamma_t
+ \varepsilon_{it}
\end{aligned}$$

### Interpretation:

- $\alpha_i$​: state fixed effects
- $\gamma_t$​: time fixed effects (e.g., macro shocks)

The progression from Pooled OLS to FE and TWFE models allows for increasingly robust estimation. The TWFE model is generally the most reliable specification.

# Results

According to all three plots, there is high statistical significance from majority of the variables within each model. Starting off with Fixed effect, all variables are statistically significant to the model. As for Two-Way FE, the variables that lack statistical significance in the Delinquency Percent, Foreclosure Rate, and Minimum Wage. Lastly, with the Pooled OLS, all variables hold statistical significance except for the constant.

Key notes to be taken about specific variables would be the $\log(DebtValue)$, which is always negative and highly significant, ranging between -1.05 and -1.21. For a 1% increase in the total debt is associated with about 1.05% - 1.21% decrease within the Debt-to-Income Ratio. As for the next significant variable, $\log(Median Household Income)$, has stronger effects within the Fixed-Effect & Pooled OLS (Ordinary Least Squares) Model. With the opposite effect of the DebtValue, it's *positive* relationship has an effect of 1% increase in Median Household Income resulting in an increase of the Debt-to-Income Ratio by \~0.75% - 1.07%.

```{r, warning=FALSE, echo=FALSE}
model_fe <- plm(
  log(DTI_ratio) ~ log(DebtValue) + log(Median_H_Inc) + MinWage +
    log(HPI) + Delinquency_pct + Med_Mpct + log(mean_CPI) +
    ForeclosureRate,
  data = model_prep,
  index = c("state", "time"),
  model = "within",
  effect = "individual"
)

model_twfe <- plm(
  log(DTI_ratio) ~ log(DebtValue) + log(Median_H_Inc) + MinWage +
    log(HPI) + Delinquency_pct + Med_Mpct + log(mean_CPI) +
    ForeclosureRate,
  data = model_prep,
  index = c("state", "time"),
  model = "within",
  effect = "twoways"
)
# stargazer(model_log_fe, type = "text")

model_pols <- plm(
  log(DTI_ratio) ~ log(DebtValue) + log(Median_H_Inc) + MinWage +
    log(HPI) + Delinquency_pct + Med_Mpct + log(mean_CPI) +
    ForeclosureRate,
  data = model_prep,
  index = c("state", "time"),
  model = "pooling"
)

stargazer(model_fe,model_twfe,model_pols, type = "text", column.labels = c("Fixed Effect","Two-Way Fixed Effect","Pooled OLS"))
# ?stargazer
```

# Discussion and Implications

## Public Policy

There are many public policies that can be implemented when it comes to the results of the data. Starting off, Income based debt relief programs are a very easy and direct form of assisting public financial health. While this only solves the problem in the Short Run, it does allow individuals to be given a second chance on debt they would've otherwise been unable to pay back. For the second option, there's the use of Financial Literacy Courses. Courses like these would have the opportunity to be implemented within any school, preferably targeted towards High-School students that are close to graduating. While this can be effective for teaching individuals who will begin facing the use of debt, it would also be effective to offer programs to individuals that are already established in their career or outside of primary schooling. This would be the most beneficial, allowing people to get practice with financial situations they may come across, and create a baseline of financial literacy to allow people to know their limits in relation to their finances and the economy overall. Lastly, a viable public policy would be to implement a sub division of the Federal Housing Finance Agency, one that would would create a financial database of understanding housing valuations via the specifications of the houses. Starting points would consist of County/State locations, House Age, House-owned Appliances retained through sale of the home, and ratings of local education.

## Business Implications

Businesses play an integral role in terms of Household Financial Health. Individuals rely on larger institutions to provide loans and the terms to be agreeable on both ends of the contract. With that said, companies are able to provide different loans, such as "Pay-Day" loans, or even "Buy-Now Pay-Later" Structures that are sold through third parties, but backed by large financial institutions. These structures are easy ways to get access to money or to even stretch payments over a period of time. The issue that presents itself here is the background checks and the usage of these financial instruments. These instruments can be used on non-essentials, and have less restrictions state-by-state in terms of capping their interest rates & debt collection. These loans aren't FDIC backed, but have large financial institutions putting forth equity on them, despite their debt collection being minimally invasive in relation to delinquent payments. This implies that the debt on these loans are utilized on the books for their high interest rates to be collected in the future. These are high risk investments due to their lack of debt collection, but the issue is significant due to it being monetarily backed by Financial Institutions. This can cause increase in rates on normal loans to offset the risk of this investment, but could cause macro-economic issues if the monetary value is large enough. Overall, the business implication from this data is to regulate these financial instruments in order to balance the incentive of utilizing backed, standardized financial instruments with lower risk.

# Conclusion

Debt-to-Income ratios are based on the premise that individuals and entities can only acquire so much debt until they become over-burdened. Companies are known for higher DTI ratios due to their ability to continuously generate profit as long as there's no significant likelihood for *Going Concern*. Some companies are more debt heavy than others, usually pertaining to specific Sectors or Industries of the market. The same relates to individuals and households, just at lower levels.

The Regression table estimates from pooled OLS, state fixed effects (FE), and two-way fixed effects (TWFE) models. Across all specifications, debt value and housing prices exhibit statistically significant negative relationships with the debt-to-income ratio, while median household income is consistently positive and significant. The magnitude of these coefficients remains relatively stable across models, suggesting robust relationships.

However, several variables—such as minimum wage, delinquency rates, and foreclosure rates—lose statistical significance in the TWFE specification. This indicates that their apparent effects in simpler models are likely driven by time-specific shocks rather than within-state variation. In particular, the coefficient on CPI changes sign when time fixed effects are included, highlighting the importance of controlling for macroeconomic trends.

Overall, the differences between the FE and TWFE models suggest that unobserved time effects play a substantial role in explaining variation in the debt-to-income ratio, and failure to account for these effects may lead to biased estimates.

# References

**Bhutta, N. (2015)**. *The ins and outs of mortgage debt during the housing boom and bust.* FEDS Working Paper 2015-41, Board of Governors of the Federal Reserve System.

**Bureau of Labor Statistics (BLS)**. (n.d.). City *CPI Data and Cumulative State CPI Data.* <https://www.bls.gov>

**Federal Housing Finance Agency (FHFA)**. (n.d.). *House price index data.* <https://www.fhfa.gov>

**Federal Reserve Economic Data (FRED)**. (n.d.). *Economic research data.* Federal Reserve Bank of St. Louis. <https://fred.stlouisfed.org>

**Federal Reserve Bank of Minneapolis**. (n.d.). \*Research and data.\* https://www.minneapolisfed.org

**Glaeser, E. L., & Gyourko, J. (2018)**. The economic implications of housing supply. *Journal of Economic Perspectives, 32*(1), 3–30.

**Mian, A., & Sufi, A. (2011)**. House prices, home equity-based borrowing, and the U.S. household leverage crisis. *American Economic Review, 101*(5), 2132–2156.

**Saiz, A. (2010)**. The geographic determinants of housing supply. *Quarterly Journal of Economics, 125*(3), 1253–1296.

# Appendix

## Usage of Median Data

Throughout the process, finding the least skewed data possible required using median values compared to the means of the data. With the scaling of incomes in the United States, the average CEO makes over 300x the average worker, slowly making it's way to 400x. The graph below shows the variance between Median & Mean Income by state.

```{r}
model_prep |> 
  group_by(state) |> 
  summarise(
    mean_income   = mean(mean_AGI,      na.rm = TRUE),
    median_income = mean(Median_H_Inc/1.25,  na.rm = TRUE),   # takes into account 1 ind. HHs
    .groups = "drop"
  ) |> 
  mutate(Income_avg_change = mean_income - median_income) |> 
  ggplot(aes(y = fct_reorder(state, Income_avg_change))) +
  geom_segment(aes(x = median_income, xend = mean_income, yend = state),
               linewidth = 1, color = "gray70", linetype = 2) +
  geom_segment(aes(x = median_income, xend = mean_income, yend = state),
               linewidth = 1, color = "gray45", linetype = 1, alpha = .65) +
  geom_point(aes(x = median_income),color = "#700000", size = 2) +
  geom_point(aes(x = mean_income),color = "green4",   size = 2) +
  labs(
    title = "Variances in Mean & Median Income (by State)",
    x = "Income",
    y = ""
  ) +
  scale_x_continuous(labels = scales::dollar) +
  theme_solarized() +
  theme(
    plot.title = element_text(family = "Times New Roman", color = "#700000", face = 'bold'),
    plot.subtitle = element_text(
      face = "italic",
      family = "Times New Roman",
      size = 10
    ),
    axis.title = element_text(family = "Times New Roman", face = "bold"),
    legend.position = "right",
    legend.title = element_text(family = "Times New Roman", face = "bold"),
    legend.text = element_text(family = "Times New Roman", face = "italic"),
    plot.caption = element_text(family = "Times New Roman", size = 7),
  )
```

## Median Income Compared to minimal Inflation 

The data here represents a non-CPI based metric of understanding data, given that the inflation rate is below-average, being 1.5%. The data here is not meant to be compared external visualizations due to it's low inflation rate. This plot is meant to display that even at a low inflation rate of 1.5%, median incomes still don't scale properly with that annual increase.

```{r}
inflation_rate <- 0.015

baseline_data <- model_prep |>
  group_by(time) |>
  summarise(avg_start_inc = mean(Median_H_Inc[time == min(time)], na.rm = TRUE)) |>
  mutate(
    years_since_start = as.numeric(difftime(time, min(time), units = "days")) / 365.25,
    inflation_line = avg_start_inc * (1 + inflation_rate)^years_since_start
  )
model_prep |> 
  mutate(state = fct_reorder2(state, time, Median_H_Inc)) |>
  ggplot(aes(x = time, y = Median_H_Inc)) + # Removed color=state from global aes
  
  # The Inflation Baseline (dashed for distinction)
  geom_line(data = baseline_data, aes(y = inflation_line), 
            color = "black", linetype = "dashed", size = .6) +
  
  # Individual State Data
  geom_line(aes(color = state), size = .5) + 
  
  facet_wrap(~state, ncol = 2) +
  
  # Your existing styling
  labs(
    y = "Median Household Income",
    x = "Time (in quarters)",
    title = "Median Household Income vs. 1.5% Inflation (Low End)",
    subtitle = "Dashed line represents standard inflation baseline from 2003",
    caption = "Vlines: 2008 Housing Crisis, 2020 COVID\nDash: 2% Inflation Baseline - www.minneapolisfed.org",
  ) +
  geom_vline(xintercept = as.Date("2008-01-01"), size = .15) +
  geom_vline(xintercept = as.Date("2020-03-01"), size = .15) +
  scale_y_continuous(labels = scales::dollar) +
  scale_color_solarized() +
  theme_solarized() +
  theme(
    legend.position = "none", # Removed as requested
    strip.background = element_rect(fill = "#eee8d5"), # Matching Solarized palette
    strip.text = element_text(family = "Times New Roman", face = "bold"),
    plot.title = element_text(family = "Times New Roman", color = "#700000", face = 'bold'),
    plot.subtitle = element_text(face = "italic", family = "Times New Roman", size = 10),
    axis.title = element_text(family = "Times New Roman", face = "bold"),
    plot.caption = element_text(family = "Times New Roman", size = 7)
  )
```
